Geometrical and physical interpretation of the action principle

### 2022-07-08 TeachingMaterial/ActionPrinciple/ActionPrincipleArticle.tex
ActionPrinciple: starting to reframe the note as an article

\begin{abstract}
We give a geometrical interpretation for the action principle in Lagrangian classical particle mechanics that also has a direct physical meaning. In a nutshell, the variation of the action between a path and its variation effectively ``counts'' the possible evolutions that ``go through'' the area enclosed. If a possible evolution is taken, all neighbouring evolutions will be parallel, making them tangent to the area enclosed by the evolution and its variation, making the action stationary. We first give the interpretation for a single degree of freedom using simple differential calculus so that it is accessible to the widest possible audience. We then generalize to multiple degrees of freedom using symplectic geometry. This treatment makes the geometry of both Hamiltonian and Lagrangian mechanics more readily evident, and gives reasons to privilege the Hamiltonian picture over the Lagrangian one.
\end{abstract}

@@

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics, which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. This allowed us to recast Hamiltonian mechanics as the dynamics of deterministic and reversible evolution, which coincidentally gives us the proper setting to characterize both the physics and the geometry underlying the principle of stationary action. We will see that the principle arises as a feature of divergence-free fields and closed two-forms, as a general mathematical property of vector calculus and differential geometry\cite{souriau1997structure}.

We will first address the case of a single degree of freedom using standard vector calculus. This provides already all the geometrical and physical insights in a language that is accessible to all physicists and engineers. Then we proceed to the case of multiple independent degrees of freedom, which requires tools from differential geometry. To make the treatment of the latter case more accessible, we will use a notation that is similar to the one used by physicists in general relativity.

>>>>

\begin{abstract}
We give a geometrical interpretation for the principle of stationary action in classical Lagrangian particle mechanics. In a nutshell, the difference of the action along a path and its variation effectively ``counts'' the possible evolutions that ``go through'' the area enclosed. If the path corresponds to a possible evolution, all neighbouring evolutions will be parallel, making them tangent to the area enclosed by the path and its variation, thus yielding a stationary action. This treatment gives a full physical account of the geometry of both Hamiltonian and Lagrangian mechanics which is founded on three assumptions: determinism and reversible evolution, independence of the degrees of freedom and equivalence between kinematics and dynamics. The logical equivalence between the three assumptions and the principle of stationary action leads to a much cleaner conceptual understanding.
\end{abstract}

@@

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics, which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. This gives a more physically motivated grounding of the classical theory and the proper setting to characterize both the physics and the geometry underlying the principle of stationary action. The assumption of determinism and reversible evolution corresponds to a diverge-free flow of the states, which preserves volumes in phase space (i.e. the count of states). The assumption of independence of the degrees of freedom means the volume reduces to the product of areas in each d.o.f., which is quantified by a closed two-form. The principle of stationary action arises as a feature of divergence-free fields and closed two-forms, as a general mathematical property of vector calculus and differential geometry\cite{souriau1997structure}. The additional assumption of equivalence between kinematics and dynamics (i.e. we can reconstruct the dynamical state simply by looking at the trajectory) is what allows us to express the principle in Lagrangian form. The argument can proceed in the reverse direction: assuming the principle of stationary action recovers a dynamical system that exhibit those three physical assumptions.

The mathematics needed to run the argument is not new, yet it is typically presented in a way and in a notation that is opaque to physicists. To provide a clear insight to the widest possible readership, we will first address the case of a single degree of freedom using standard vector calculus. This provides already all the geometrical and physical insights in a language that is accessible to all physicists and engineers. Then we proceed to the case of multiple independent degrees of freedom, which requires tools from differential geometry. Here we will use a notation that is similar to the one used in general relativity. We will concentrate mainly on the geometrical and physical meaning of the mathematical objects and their relationships, leaving the full argument and calculations in the appendix.
